Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
put Overall Algorithm after Architectural Details

\section{Extension to Neural Networks}
EAGGA as implemented by \citet{EAGGA} applies the optimizer to XGBoost, which proved to vastly outperform a union of competitor models with respect to
dominated hypervolume and achieved comparable or better performance than ParEGO optimizing over the
extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}.
Our work extends EAGGA to neural networks. We chose this model class for several reasons:
\begin{enumerate}[label*=\arabic*.]
  \item NNs are notoriously uninterpretable due to complex transformations of the feature spaces
  \item Non-linear activation functions implicitly model interactions without direct control of which features interact with each other
  \item There are no monotonicty guarantees
\end{enumerate}

\subsection{Overall Algorithm}
We implement EAGGA largely as described by \citet{EAGGA} with only minor changes due to no straightforward transferability either due to different implementations
of some classes and functions in R compared to Python or differences in model properties between XGBoost and neural nets.\\
% TODO: after architectural details:
Our NNs use the following hyperparameters initialised as described
\begin{itemize}
  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$.
  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$.
  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15.
\end{itemize}
For the group structure initialization we also employ a feature, interaction, and monotonicity detector.
\\
The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features used as
probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
target. In these edge cases we use all features with non-zero filter-values instead.
\\
The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm computing the interaction scores we don't use the residual
sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
\\
The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
\\
Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
the remaining 20\% to compute their scores.

@@

Grouping the network into sub-networks enables restricting feature interactions to only those features included in the respective group.

@@

than the multiplicative one happening, for instance, in a linear model.

@@

@@

NI sums the numbers of all possible pairwise interactions in each group $\sum_g^G {p_g \choose 2}$ and divides them by the number of all possible pairwise interactions

>>>>

\section{Extension to Neural Networks}
EAGGA as implemented by \citet{EAGGA} applies the optimizer to XGBoost, which proved to vastly outperform a union of competitor models with respect to
dominated hypervolume and achieved comparable or better performance than ParEGO optimizing over the
extended search space $\check{\boldsymbol\Lambda}$ \citep[pp. 543-545]{EAGGA}.
Our work extends EAGGA to neural networks. We chose this model class for several reasons:
\begin{enumerate}[label*=\arabic*.]
  \item NNs are notoriously uninterpretable due to complex transformations of the feature space
  \item Non-linear activation functions implicitly model interactions without direct control of which features interact with each other
  \item There are no monotonicty guarantees in neural networks
\end{enumerate}
The optimizer is implemented largely as described by \citet{EAGGA} with only minor changes due to there being no straightforward transferability of some functions
from R to Python or differences in model properties between XGBoost and neural nets.\\

@@

Grouping the network into sub-networks enables restricting feature interactions to only those features that are included in the respective group.

@@

than the multiplicative one known, for instance, from the linear model.

@@

\subsection{Overall Algorithm}
Our NNs use the following hyperparameters initialised as described
\begin{itemize}
  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$
  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$
  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15
\end{itemize}
For the group structure initialization we also employ feature, interaction, and monotonicity detectors.
\\
The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features as
probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
target. In these edge cases we use all features with non-zero filter-values instead.
\\
The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm, when computing the interaction scores we don't
use the residual sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
\\
The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
\\
Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
the remaining 20\% to compute their scores.

@@

NI sums the numbers of all possible pairwise interactions in each group ${p_k \choose 2}$ and divides them by the number of all possible pairwise interactions
